---
title: "Data Structure in R"
author: "Dr. Maïwen Caudron-Herger, PD Dr. Carl Herrmann"
date: "10/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Description of how data is organized and displayed.

# 1 - Simple variables

```{r}
a = 2
x <- 4
# to assign values to variables, <- and = are equivalent
b = TRUE
# or
b <- TRUE
name = 'Rosalinde'
```

We can perform tests on simple variable using the `==`, `>`,`<` operators

```{r}
name == 'rosalinde'
name == 'Rosalinde'
a > 5
```


# 2 - Vectors

The function **c()** combines vectors of length 1 into longer vectors. The number of elements in a vector can be determined with **length()**.

```{r}
x <- 9
length(x)	            # [1] 1
                      # x is a vector of length 1.
#
y <- c(x, 3, 7, 2)
length(y)	            # [1] 4
                      # y is a vector of length 4.
```

Vectors can be made of numbers, strings and characters or logicals (TRUE, FALSE, NA = not available).

```{r}
z <- c("a", "b", "abc")
length(z)	                        # [1] 3
w <- c(TRUE, FALSE, TRUE, TRUE)
str(w)	                          # logi [1:4] TRUE FALSE TRUE TRUE
```

Note the difference between logical and string

```{r}
w <- c("TRUE", "FALSE", "TRUE", "TRUE")
str(w)	                          # chr [1:4] "TRUE" "FALSE" "TRUE" "TRUE"
```

When the data types are mixed, all elements of the vector are converted into the **lowest** type.

```{r}
w <- c(TRUE, "FALSE", 3, 1+0i)
str(TRUE)	                        # logi TRUE
str("FALSE")	                    # chr "FALSE"
str(3)	                          # num 3
str(1+0i)	                        # cplx 1+0i
str(w)  	                        # chr [1:4] "TRUE" "FALSE" "3" "1+0i"
```

To select elements of a vector, the name of the vector is followed by brackets **[]** including the index of the element.

```{r}
y[3]	                    # [1] 7
y[2:3]	                  # [1] 3 7
y[3:2]	                  # [1] 7 3
y[c(2,4)]	                # [1] 3 2
y <- y[-2]	              # [1] 9 7 2
y[-2]                     # Does not change the y vector 
y[y < 5]	                # [1] 3 2 -> gives you the elements of y with value < 5!
```

We can also perform **tests** on elements of a vector.  
These tests will be applied for each element:  

```{r}
x = c(1,4,3,7)
x < 4
values = c('high','high','low','high')
values == 'high'
```

The function **which()** can be extremely useful

```{r}
y[which(y < 5)]          # 
# [1] 2 4 -> gives you the indexes/positions of the elements in y which are < 5!
```

You can modify elements in a vector:

```{r}
y	                        # [1] 9  3  7  2
y[1] <- 10
y	                        # [1] 10  3  7  2
y[] <- 8
y	                        # [1] 8 8 8 8
y <- 8
y	                        # [1] 8
```

It is possible to perform operations with the logical TRUE and FALSE:

```{r}
values = c('high','high','low','high')
values == 'high'
eval <- (values == 'high')
sum(eval)
# Can you explain/understand why sum(eval) does return 3?
```



# 3 - Matrices

A matrix is an **array** of numbers. They are defined using the function matrix as followed:

```{r}
y <- c(1, 2, 3, 1, 2, 3)

X <- matrix(data = y, nrow = 2, ncol = 3, byrow = TRUE)
X
#       [,1] [,2] [,3]
# [1,]    1    2    3
# [2,]    1    2    3
X <- matrix(data = y, nrow = 2, ncol = 3, byrow = FALSE)
X
#
X <- matrix(data = c(1,2,3,4,5,6), nrow = 2, ncol = 3, byrow = TRUE)
X
#       [,1] [,2] [,3]
# [1,]    1    2    3
# [2,]    4    5    6

```

```{r Exercise 1}
## Exercise 1:
# Create a matrix that look like:
#       [,1] [,2] [,3]
# [1,]    2    4    6
# [2,]    8    10   12
# [3,]    3    5    7
# [4,]    2    9    11
```

The parameter **byrow** is per default set to **FALSE** and the matrix is constructed column-wise.

```{r}
X <- matrix(data = y, nrow = 2, ncol = 3)
X
#       [,1] [,2] [,3]
# [1,]    1    3    2
# [2,]    2    1    3
```

To access the elements of a matrix, it is similar to the vector but with 2 indexes.

```{r}
X[1, 2]	                    # -> first row and second column
                            # [1] 3
X[1:2, 1]	                  # -> first and second rows, first column
                            # [1] 1 2	
                            # is vector of length 2
length(X[1:2, 1])	          # [1] 2
```

It is possible to remain in 2 dimensions.
You can ask for the **dimension** of a matrix (and see later also for dataframe) using the function **dim()**.

```{r}
X[1:2, 1, drop = FALSE]
#      [,1]
# [1,]    1
# [2,]    2

dim(X[1:2, 1, drop = FALSE])
```

```{r Exercise 2}
## Exercise 2:
# Select part of the matrix: for example
X[c(1,2),c(1,2)]        # Same as X[1:2,1:2]
# Try something else
```

We can also build a matrix by joining multiple column-vectors using `cbind`:

```{r}
## create 3 vectors of 10 numbers each
x = rnorm(10)
y = rnorm(10)
z = rnorm(10)

## now join these 3 columns into a matrix
A = cbind(x,y,z)
```

Dimension of a matrix/data.frame:

```{r}
dim(A)
```

column/row names

```{r}
A
```

```{r}
rownames(A) = letters[1:10]
colnames(A) = c('patient1','patient2','patient3')
A
```

transposing a matrix/data.frame

```{r}
A
```

```{r}
t(A)
```



# 4 - Arrays

**Remember**

> -   Vectors have dimension n = 1 (1D)
> -   Matrices have dimension n = 2 (2D)
> -   Arrays can have n dimensions (nD)

```{r}
X <- array(1:40, dim = c(5, 4, 2))
X
# , , 1
#
#        [,1] [,2] [,3] [,4]
# [1,]    1    6   11   16
# [2,]    2    7   12   17
# [3,]    3    8   13   18
# [4,]    4    9   14   19
# [5,]    5   10   15   20
#
#, , 2
#
#         [,1] [,2] [,3] [,4]
# [1,]   21   26   31   36
# [2,]   22   27   32   37
# [3,]   23   28   33   38
# [4,]   24   29   34   39
# [5,]   25   30   35   40
```

To access the elements of an array, it is similar to the matrices but with 3 indexes.

```{r}
X[2, 3, 1] # second row, third column, first level
#
X[4,3,2]
```

**Question**: when could it be useful?

# 5 - Lists

Lists are very flexible structures and can contain **various types of elements**.\
Elements of a list can be named - they do not have to be named but it can be helpful.

```{r}
List1 <- list(color = c("blue", "red"), size = 5, state = TRUE)
List1
# $color
# [1] "blue" "red"

# $size
# [1] 5

# $state
# [1] TRUE

list3 <- list(cold_stuff = c("butter", "milk"), warm_stuff = c("cake", "Meat"), beaker = "bread", number = 10, supermaket = TRUE)
list3
```

We can inspect this list and access the elements using the "**\$**" sign and their name.

```{r}
str(List1$color)              # chr [1:2] "blue" "red"
List1$color                   # [1] "blue" "red" 
List1[["color"]]              # [1] "blue" "red"
                              # This is of length = 2
# but 
List1["color"]                # $color
                              # [1] "blue" "red"
                              # This is if length = 1
```

So if you try to access the elements, there is a difference:

```{r}
List1[["color"]][1]
List1["color"][1]
```

```{r Exercise 3}
## Exercise 3:
# What would happen if you type List1["color"][2] ?
# Try it!
```

How does it look for a list without named elements?

```{r}
List2 <- list(c("blue", "red"), 5, TRUE)
List2
# [[1]]
# [1] "blue" "red" 

# [[2]]
# [1] 5

# [[3]]
# [1] TRUE

List2[1]	                  # First element of List2
                            # [[1]]
                            # [1] "blue" "red"

List2[[1]]	                # Content of the first element of List2
                            # [1] "blue" "red" 

List2[[1]][2]	              # Second position of the first element of List2
                            # [1] "red"
```

```{r Exercise 4}
## Exercise 4:
# Create a list with a first element of length 1, a second element of length 3 and a third element of length 5.
# Name the elements of the list.
# How would you access the 4th position of the third element of your list?

# How to attribute names to the elements of List2?
# You can monitor the changes made to your list in the RStudio Environment window.
# Click on the arrow in the Environment window:
names(List2) <- c("color", "size", "state")
```

**unname(List2)** returns the unnamed List2, but does not change List2.

```{r}
# To unname List2:
List2 <- unname(List2)
```

How to remove an element from a list?

```{r}
List2[[3]] <- NULL
```

How to remove a position from an element of the list?

```{r}
# Replace the element by itself (knowing it is a vector) without the position you want to remove:
List2[[1]]
# [1] "blue" "red" 
List2[[1]][-1]
List2[[1]] <- List2[[1]][-1]       # remove position 1
List2[[1]]
# [1] "red"
```

```{r Exercise 5}
## Exercise 5:
# Add a second and third position in the first element of List2.
```

# 6 - Dataframes

Dataframes can be seen as special lists of vectors of the same length.  
Dataframes can be indexed similar to lists and matrices.  

Data frames are like matrices, but they can contain multiple types of values mixed.  
We can create a data frame using the `data.frame` command. We can also convert a matrix into a data frame with `as.data.frame`.  
However, there are important differences between data frames and matrices!

```{r Exercise 6}
## Exercise 6:
# Create a matrix C with 4 rows and 5 columns with numbers from 1 to 20
# Create a data frame D using the matrix C
# Now, execute the following:
# C[2,3]
# C[4]
# D[2,3]
# D[4]
# Do you see and understand the difference?
```



Prepare 3 vectors of the same length - 4 elements for example.

```{r}
Name <- c("Ben", "Alice", "Daniel", "Charlie")
Year <- c(1990,1992,1995,1993)
Option <- c("Biology","Physics", "Chemistry","Mathematics")
```

Create a dataframe using the function **data.frame()**

```{r}
Students <- data.frame(First_Name = Name, Year = Year, Options = Option)
Students
# 		First_Name	Year	Options
# 	1	Alice	      1990	Biology
# 	2	Ben	        1992	Physics
# 	3	Charlie	    1995	Chemistry
# 	4	Daniel	    1993	Mathematics
```

To access the elements of the dataframe, use the sign "**$**" and indexes "**[]**".

```{r}
Students$Year
                  # [1] 1990 1992 1995 1993
                  # is similar to Students[["Year"]]
Students[["Year"]]

length(Students$Year)
                  # [1] 4
Students$Options[3]
                  # [1] Chemistry
Students[2:3,3]
                  # [1] Physics   Chemistry
```

Dataframes can be **merged**  
For example, create a second dataframe, with partial info from the "Students" dataframe:

```{r}
Full_Names <- data.frame(First_Name = c("Alice","Ben","Daniel"),
                         Last_Name = c("Parker","James","Brown"))
merge(Students, Full_Names)
#  		First_Name 	Year	Options	    Last_Name
#	  1	Alice 	    1992	Physics    	Parker
#	  2	Ben 	      1990	Biology     James
#	  3	Daniel 	    1995 	Chemistry   Brown
```

By defaults, **all** is set to "**FALSE**" and only elements with entries in both dataframes are merged.  
Help on the function "merge" can be obtained with **?merge**.  
```{r}
merge(Students, Full_Names, all=TRUE)
Students <- merge(Students, Full_Names, all=TRUE)
```


### Further manipulation of a dataframe:

> -   Addition of a new column
> -   Delete a column or a row
> -   Changing the name of the columns/rows
> -   Combine dataframe column/row wise
> -   Operations on numerical dataframes
> -   ...


# 7 -  Loops

Sometimes, we need to repeat the same operation of all elements of a vector, or all rows/columns of a matrix or data.frame. This is a **loop**.


## Loops over elements of a vector: the `sapply` function

We can perform a loop over all elements of a vector using the `sapply` function: `sapply` need 2 informations : (1) which vector do we consider? (2) which function do we want to apply to each element of this vector?

```{r}
## create a numerical vector
x = rnorm(20)
x
```

For each element of the vector, determine if it is positive:

```{r}
sapply(x,function(y) {y>0})
```

What happens here is that during the loop, the variable `y` will take all values from the `x` vector one by one, and the corresponding function will be applied to this element.

We could have obtained the same result here with: 

```{r}
x > 0
```

However, we can do more sophisticated things here: let us introduce the `ifelse` function:

```{r}
ifelse(2>5,'This is true','This is not true')
```

See? a test is performed; if the test returns `TRUE`, the first operation is executed, if `FALSE`, the second operation is executed.

Let us use `ifelse` in a `sapply` function:

```{r}
sapply(x,function(y) { ifelse(y>0,y^2,0)})
```


## Loops over row/columns of a matrix: the `apply`-function

Using the `apply` function, we can repeat a certain task over all row or columns of a matrix:

```{r}
## determine the mean over all rows
apply(A,1,mean)
```

We can combine this to find, for example, which rows have a mean higher than 0.5:

```{r}
which(apply(A,1,mean) > 0.5)
```

Similarly, we can repeat a task over all columns of a matrix:

```{r}
apply(A,2,max)
```

We can also use the `sapply` function to perform a loop over the columns of a data frame (similar to the `apply` function):

```{r}
## create a data frame
Dat = data.frame(c(1,2,3,4,5,6,7,8,9,10), c(rnorm(10, 15, 4)), c(rpois(10, 3)), c(rbinom(10, 1, 0.4)))
colnames(Dat) = c("X", "Variable A", "Variable B", "Variable C")
Dat
```

Now compute the mean over all columns using the `sapply` function:

```{r}
sapply(Dat,mean)
```

BEWARE! This does not work for a **matrix**:

```{r}
Mat = as.matrix(Dat)
sapply(Mat,mean)
```

What just happened?

# SUMMARY SECTION

### What have we learned today?

> -   Introduction to R Markdown
> -   Where can I get help?
> -   Introduction to RStudio and R
> -   Assignment instruction
> -   Delete a variable
> -   Load RData file
> -   print() and cat() functions
> -   Data structure in R
> -   Vectors
> -   Matrices
> -   Arrays
> -   Lists
> -   Dataframes
> -   Associated important functions such as apply and sapply


# <span style="color:#9F2042">HOMEWORK ;-)</span>

<span style="color:#9F2042">Please have a look at the two **cheatsheets** about R:</span>  
<span style="color:#9F2042">- R command cheatsheet.pdf</span>  
<span style="color:#9F2042">- base-r.pdf</span>
<span style="color:#9F2042">- rmarkdown-reference.pdf</span>

<span style="color:#9F2042">Think about examples to be added to this file.</span>  

------------------------------------------------------------------------  
