---
title: "Exercise Sheet 3 - Hierarchical Clustering / PCA"
author: "Carl Herrmann - Maiwen Caudron-Herger"
output: html_document
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print="120")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=120)

```

We are going to work with a dataset of gene expression of patient with leukemia. There are 2 types of leukemia: ALL (acute lymphoid leukemia) and AML (acute myeloid leukemia).
We have
* one file with gene expression
* one file with clinical annotatations for the patients

Let's read the data for this tutorial:

```{r, echo=FALSE, eval = TRUE}

#### This is all pre-processing of the all/aml dataset
#### RUN THIS CHUNK AT THE BEGINNING

all.aml.anno = readRDS(url('https://www.dropbox.com/scl/fi/kfnxs5ltldmfj6px1sp13/all.aml.anno.rds?rlkey=yhahfyk4h240fbt1vctt3bp5b&dl=1'))
all.aml.exp = readRDS(url("https://www.dropbox.com/scl/fi/grt6yud9yjmwqx3zgwhte/all.aml.exp.rds?rlkey=hs8p83tkxvy27t8v38tw51x7i&dl=1"))

```

# 0. Recap of the last sheet

In the last sheet we have kept working with our diabetes dataset by learning how to **measure centrality** and other statistical information such as **quantiles**. Furthermore, we have learned functions to measure correlation between variables and explored different methods to do so. Finally, on a new dataset, we have moved over to the concept of unsupervised learning and used one of its most simple but effective methods, k-means clustering.


# 1. Objectives of this week

In this sheet we will conclude the topic of **unsupervised learning** by looking at **hierarchical clustering** and **Principal Component Analysis (PCA)**.

# 2. Hierarchical clustering

Clustering is a method by which we group together similar observations while separating out the dissimilar ones. We will cluster our samples from the cancer dataset to see which samples cluster together or separately. Hierarchical clustering does not generate discrete clusters of datapoints, but rather creates a dendrogram that indicates the magnitude of similitude between samples. Once again is up to the Data Scientist to decide the right amount of clusters.

### Using the most variable, thus informative genes

When performing clustering, we usually reduce the number of genes used, as some of them are not informative. For example, genes that show a mostly constant expression across all samples will not be useful to distinguish the samples, right? One simple method is to select genes that show a **high variance** across all samples.

```{r}
## compute the variance for all genes accross all samples
topVar = apply(all.aml.exp, 1, var)
summary(topVar)
```

We now want to find the top 25% with the highest variance

```{r}
## what is the 75% quantile of the variance?
q75 = quantile(topVar, probs = 0.75)
q75
```

So let us select all rows (genes) with a variance higher than `q75`:

```{r}
i.topvar = which(topVar >= q75)

## only select the gene with a variance in the top 25%
all.aml.exp.topVar = all.aml.exp[i.topvar,]
dim(all.aml.exp.topVar)
```


### Computing the correlation between all patients (columns)

```{r}
## Creating a correlation based distance matrix
cor.mat = cor(all.aml.exp.topVar, method="pearson")
```

Let us display the correlation matrix as a heatmap, using the `pheatmap` function:

```{r}
library(pheatmap)
pheatmap(cor.mat)
```

Each cell of this heatmap represents the correlation value between the sample in the row and the sample in the column. The correlation of a sample to itself is always 1 (red diagonal).

The function automatically determines the clustering trees of the rows and columns (which are identical, since the correlation matrix is symmetrical!)

Note the color scale: it is biased by the 1's in the diagonal. Since these values are trivial, we can remove them to avoid distorting the color scale:

```{r}
diag(cor.mat) <- NA
```

```{r}
pheatmap(cor.mat)
```


### Plotting the dendrogram along with clinical annotations

This is a nice representation, but in order to interpret this clustering, we need to add some additional (clinical) information to interpret the clustering structure. To do this, we will create an annotation data frame containing as columns a number of clinical features.

The clinical annotation is stored in the `all.aml.anno` data frame.

```{r}
head(all.aml.anno)
```


We can now plot again the heatmap, using the annotation dataframe to add additional information

```{r,fig.height=8,fig.width=10}
pheatmap(cor.mat,annotation_row = all.aml.anno,annotation_col = all.aml.anno)
```

> How would you interpret this dendrogram? How many clusters do you observe? Do they make clinical sense?

> The function `pheatmap` accepts a parameter clustering_method to indicate alternative linkage methods; try other linkage methods (check which are available with the pheatmap help page!)

> can you redo this analysis using all the genes instead of the most variable ones? Do you see a difference?

# 3. Dimensionality reduction - PCA

We will now use principal component analysis to explore the same dataset, and identify directions (i.e. principal components) with maximal variance. Principal components Analysis finds n-dimensional vectors (Principal Components) in the direction of the largest variance, thereby allowing you to describe an n-dimensional dataset with just a few dimensions. Think back of how you could describe most of the distribution from exercise set B with just one vector, even though the data was 2-dimensional.

```{r}
pca = prcomp(t(all.aml.exp.topVar), center = F, scale. = F) # We set these as false as we have already scaled our data
print(pca)
```

> How many principal components do you obtain? Compare this to the dimension of the matrix using the `dim` function! What would happen if you would not transpose the matrix with t(...) in the prcomp function?

Principal components are ranked by the amount of variance that they explain. This can be visualized using a **scree plot**, indicating how much variance each PC explains: the proportion of **standard deviation** explained by each principal component is contained in the `pca$sdev` vector:

```{r}
pca$sdev
```

We see that the standard deviation is indeed going down! Let us now plot the proportion of total variance explained by each PC

```{r}
variance = (pca$sdev)^2
prop.variance = variance/sum(variance)
names(prop.variance) = 1:length(prop.variance)
barplot(prop.variance[1:20],ylab='Proportion of variance') # we only plot the first 20 PCs
```

We can now display the data points (i.e. patients) in the first two principal components. In addition, we can color the dots according to certain clinical parameters:

```{r}
colors = c('blue','green','red','magenta')

plot(pca$x[,1], pca$x[,2], 
     col= colors[all.aml.anno$ALL.AML], pch=19,
     xlab='PC1',ylab='PC2') 
legend('topright',legend=levels(all.aml.anno$ALL.AML),fill=colors)
```

> Color the patients according to the distinction between bone marrow samples and peripheral blood, then redo the plot by including some other PCs (e.g. PC1 vs. PC3)
> Color the patients according to the distinction between genders (fourth column) then redo the plot by including some other PCs (e.g. PC1 vs. PC3). Are there any groups? If there were, what could this mean?



#### Exercice 1 - Hierarchical clustering and kmeans

1. Based on the heatmap you have obtained from the correlation matrix, perform a kmeans clustering of your patients with the corresponding number of clusters (BEWARE that you need to transpose the matrix first!)
`km <- kmeans(t(all.aml.exp), centers = k)`
2. Add the cluster membership to the all.aml.anno dataframe as a new column cluster
3. Redo the heatmap: the cluster membership should appear
4. Repeat with other values of k


#### Exercise 2 - Hierarchical clustering

In section 2, we have computed a correlation matrix, and used this matrix to build a clustering tree.

1.  Create a 4x4 distance matrix, following the example below (but with other values!); beware that a distance matrix should be **symmetrical**!!

```{r}
## this is the example from the lecture slides
d = matrix(c(0,2,3,5,2,0,1,4.1,3,1,0,4,5,4.1,4,0),nrow=4)
rownames(d) = letters[1:4]
colnames(d) = letters[1:4]
pheatmap(d)
```

Try different linkage methods using the `clustering_method` parameter to see if the topology of the dendrogram changes!

2.  Try building a distance matrix which would lead to different topologies of the dendrogram, depending on which linkage method is used! Show the dendrograms built with different linkage methods!



#### Going further *(expert)*

Instead of performing the k-means on the whole gene expression matrix, we can run k-means on the space in which each patient is represented by the first k principal components (available in `pca$x`)

1.  Run k-means with different numbers of clusters (1-10) on the patients using the first 2, 4, 6,... principal components (i.e. the first columns of `pca$x`). Use the elbow method to evaluate how the within sum of squares (WSS) evolves. What is the optimal number of clusters? *Hint*: run a `sapply` loop with the `kmeans()` function and choose the WSS from the `$tot.withinss` attribute (you can refer to the end of last week's exercise sheet 02 for an example). Plot the WSS using the `plot()` function.

2.  Represent the patients in the PCA plot as previously, but color them according to the cluster they belong to! Run kmeans with two clusters for this and choose the cluster assignment from the `$cluster` attribute.

