---
title: "Exercise Sheet 06: hypothesis testing"
author: "Carl Herrmann, Maïwen Caudron-Herger"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print="120")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=120)
opts_knit$set(root.dir  = "~/")
```

# 0. Recap of the previous sheet

In the last two exercises sheets you learnt about different major types of probability distributions like Normal, Binomial, Negative binomial, Student t etc.
You have also gained pratical experience with the central limit theorem and the confidence intervals.

# 1. Introduction and objectives

In this exercises sheet you will start to see how to formulate hypotheses and how to test them.
We want to learn how to use and interpret hypothesis tests.
We will mainly focus on test on **mean values**

# 2. First statistical test

We will work again on the ALL/AML expression dataset that we worked on for the exploratory data analysis.
We cleaned up the dataset to remove extreme values, and to include gene symbols for better interpretation of the results.

```{r}
all.aml = read.delim('http://bioinfo.ipmb.uni-heidelberg.de/crg/datascience3fs/practicals/data/all.aml.cleaned.csv',header=TRUE)
all.aml.anno = read.delim("http://bioinfo.ipmb.uni-heidelberg.de/crg/datascience3fs/practicals/data/all.aml.anno.cleaned.csv", header=TRUE)
#
# We convert all.aml into a data matrix rather than a data.frame
all.aml = data.matrix(all.aml)
```

We first check how the values are distributed for a gene across all the samples (= patients).
For that, we take a random gene, and plot the histogram of the values across the samples.
You should run this code several times, as at each run, another random gene is chosen!

```{r}
# Sample a random line from the expression table
i = sample(1:nrow(all.aml),1)
#
# Plot the histogram
hist(all.aml[i,],
     main=rownames(all.aml)[i],
     col='lightblue',
     breaks=20, 
     xlab = "Expression level")
```

We can now verify the **normality** of the distribution using a QQ-plot

```{r}
# Sample a random line from the expression table
i = sample(1:nrow(all.aml),1)
#
# Plot the histogram
qqnorm(all.aml[i,],main=rownames(all.aml)[i]); qqline(all.aml[i,])
```

If the data is normally distributed, the dots representing the quantiles should lie on the straight line.
Here again, run the code several times, and check the QQ-plot for several random genes!

> Search for the line corresponding to the gene TP53 using the `grep` function (check the help!) inside the vector of row names `rownames(all.aml)` and check the normality of the distribution for this important oncogene.

```{r}
# Get the line for the gene TP53 using the function grep()
i.tp53 <- grep("TP53",row.names(all.aml))
# line 438
#
# Check the normality of the distribution
qqnorm(all.aml[i.tp53,],main=rownames(all.aml)[i.tp53]); qqline(all.aml[i.tp53,])
```

> The QQ-plot is a usefull tool to visually inspect the normality.
> However, it does not give a quantitative measure of the normality!\
> We will see later on how we can determine in a more quantitative manner the normality of the data.

------------------------------------------------------------------------

#### Exercise set A

1.  Select a gene randomly and calculate the mean expression and standard deviation (sigma) for the first and second half of the patients separately.

2.  Do the means lie within a 3 sigma interval (two sided) from each other?
    Or outside?

------------------------------------------------------------------------

Now, we want to use a statistical test to check if, for a given gene,

-   there is a significant difference in the mean expression of this gene between two groups (for example ALL vs AML, 2-sample test); or
-   this gene has an expression which is significantly above a certain value (1-sample test)

This is exactly what **mean tests** such as the t-test (or the Wilcoxon tests) are designed for!
We will perform here a t-test.

## Two-sample t-test

We will now compare the expression of certain genes between ALL and AML patients.
For a gene X, the question is:

*Is there a significance difference in expression of gene X between ALL and AML patients?*

> Formulate the $H_0$ hypothesis in this case!
> Take the gene EIF4B as an example, an let us check if there is a significant different in gene expression between the 2 groups.

```{r}
# Get the column numbers of the ALL patients
i.all = grep('ALL', all.aml.anno$ALL.AML)
#
# Find the line number of the gene EIF4B
i.gene = grep('EIF4B',rownames(all.aml))
#
# Now, let us plot the expression in the 2 groups
boxplot(list(ALL=all.aml[i.gene,i.all],AML=all.aml[i.gene,-i.all]),
        main=rownames(all.aml)[i.gene])
```

> Does that look like a big expression difference?
> Redo this analysis for the gene CCND3 and OS9.
> What is your impression?

```{r}
# Find the line number of the gene CCND3
i.gene = grep('CCND3',rownames(all.aml))
#
# Now, let us plot the expression in the 2 groups
boxplot(list(ALL=all.aml[i.gene,i.all],AML=all.aml[i.gene,-i.all]),main=rownames(all.aml)[i.gene])
```

```{r}
# Find the line number of the gene OS9
i.gene = grep('OS9',rownames(all.aml))
#
# Now, let us plot the expression in the 2 groups
boxplot(list(ALL=all.aml[i.gene,i.all],AML=all.aml[i.gene,-i.all]),main=rownames(all.aml)[i.gene])
```

> Now, let us perform a statistical test to check if the H0 hypothesis for each of these 3 genes is valid or not.

-   Compute the t-value for a **2-sample** t.test for each of the genes

```{r}
# Find the line number of the gene EIF4B
i.gene = grep('EIF4B',rownames(all.aml))
#
# Compute the t-value. Check the help for the function t.test().
t.obs = t.test(all.aml[i.gene,i.all],all.aml[i.gene,-i.all])$statistic
t.obs
```

**Is this a large value (in absolute)?** What value would be typically obtained if there were no difference (that is, if $H_0$ is true)?
Let's have a look at this below!

## One-sample t-test

Using a one-sample t-test, we can check if values differ significantly from a target value.
For example, you could sample 10 chocolate bars, and test if they significantly differ from the expected weight of 100g:

```{r}
bars = c(103,103,97,102.5,100.5,103,101.3,99.5,101,104)
t.test(bars,mu = 100)
```

> How would you interpret this result?
> Should we perform a one- or two-tailed test?

**Beware not to get confused between one-/two-sample tests, and one-/two-tailed tests!!!**

Here, we would like to check which genes are significantly more expressed than the average expression; hence, we will

1.  for each gene, compute the average expression over all patients
2.  compute the average of these average values (now over all genes), and obtain one number (e0).
3.  we will perform a one-sample test over all genes, and compare their expression with the value e0.

```{r}
e0 = mean(apply(all.aml,1,mean))
e0
```

------------------------------------------------------------------------

#### Exercise set B

1.  Perform a one-sample t-test for all genes, and test for significant deviation from the value e0.

2.  Extract the list of significant genes!

------------------------------------------------------------------------

<<<<<<< Updated upstream:06_HypothesisTesting.Rmd
# 3. Checking the normality of the distribution
=======
### Exercice 3

Can you find interesting differences in the mean values of parameters of the dataset **dat** for the two groups defined by the **location**?
Follow these steps:

1.  Select the two groups according to the location.
    To do so, check the result of `unique(dat$location)` and create two vectors that contain the corresponding row indexes of "dat".

2.  Calculate the mean values of the numerical parameters for each group (ex: age, height, weight, waist, hip, ...).
    *Hint*: create first a dataframe with numeric columns only (subset the current data frame "dat" using c(2,3,4,5,6,8,10,11,13,14,17,18)).
    Select the rows corresponding to the two groups and use an `sapply()` loop to calculate the mean values for each.

3.  Select one of these, formulate the H0 and H1 hypotheses and perform a (two-sided) t-test.
    Interpret the result (alpha = 0.05).

## ------------------------------------------------------------------------

# 3. Going further ... Checking the normality of the distribution

In principle, t-tests require that the data is approximately normally distributed.
If not, we can use **non-parametric** tests (see next lecture).
>>>>>>> Stashed changes:archive/06_HypothesisTesting.Rmd

In order to check whether the data is normally distributed or not, it is possible to perform a **Shapiro-Wilk** normality test (see lecture).
This statistical test is implemented in R in the function **shapiro.test()**.

Let's have a look at the normality of the genes in our dataset all.aml.

First, we perform a Shapiro-Wilk test on the gene TP53, which we already tested before, using a QQ-plot.

```{r}
i.tp53 <- grep("TP53",row.names(all.aml))
# line 438
#
# Check the normality of the distribution using a QQ-plot (Reminder)
qqnorm(all.aml[i.tp53,],main=rownames(all.aml)[i.tp53]); qqline(all.aml[i.tp53,])
#
# Perform a Shapiro-Wilk test
shapiro.test(all.aml[i.tp53,])
# extract the p-value
shapiro.test(all.aml[i.tp53,])$p.value
# Reminder: for p-value >= 0.05, the data is normally distributed. For p-value < 0.05, the data is not normally distributed.
```

With a p-value = 0.2933688, the data for TP53 is normally distributed!

> Determine which genes in our all.aml dataset are or not normally distributed.
> Select a gene that is normally distributed according to the Shapiro-Wilk test and visualize its distribution using **both** a histogram and a QQ-plot.
> Can you find a way to fit the histogram with a normal distribution?
> (use the function dnorm or rnorm - determine the parameters you need first) Bonus question: how can I randomly obtain a QQ-plot for the genes which are normally distributed?

We found that the gene MAPK3 and RING1 were normally distributed.
Here are the gene expression values for these two genes:

```{r}
i.mapk3 <- grep("MAPK3",row.names(all.aml))
hist(all.aml[i.mapk3,])
qqnorm(all.aml[i.mapk3,],main=rownames(all.aml)[i.mapk3]); qqline(all.aml[i.mapk3,])
#
i.ring1 <- grep("RING1",row.names(all.aml))
hist(all.aml[i.ring1,])
qqnorm(all.aml[i.ring1,],main=rownames(all.aml)[i.ring1]); qqline(all.aml[i.ring1,])
```

Now that we know that the gene expression values are normally distributed, we can perform a statistical test and question whether the two genes are differentially expressed between ALL and AML patients (this we've done above already) or whether there is a difference between male and females etc ...

```{r}
# Check the annotation dataset
head(all.aml.anno)
# Let's find some differences
# We already have i.mapk3 and i.ring1
i.f <- grep('M', all.aml.anno$Gender)
i.bm <- grep('BM', all.aml.anno$BM.PB)
# Case 1:
boxplot(list(Female=all.aml[i.mapk3,i.f],Male=all.aml[i.mapk3,-i.f]),main=rownames(all.aml)[i.mapk3])
#
# Case 2:
boxplot(list(Female=all.aml[i.ring1,i.f],Male=all.aml[i.ring1,-i.f]),main=rownames(all.aml)[i.ring1])
#
# Case 3:
boxplot(list(BoneM=all.aml[i.mapk3,i.bm],P_Blood=all.aml[i.mapk3,-i.bm]),main=rownames(all.aml)[i.mapk3])
#
# Case 4:
boxplot(list(BoneM=all.aml[i.ring1,i.bm],P_Blood=all.aml[i.ring1,-i.bm]),main=rownames(all.aml)[i.ring1])
```

Let's consider the 1st case and check the mean values.

```{r}
mean(all.aml[i.mapk3,i.f])
mean(all.aml[i.mapk3,-i.f])
```

Let's take the H0 hypothesis:\
"The average expression levels of MAPK3 in female patients is equal to the average expression levels male patients."\
\> What kind of t.test are we going to perform?

```{r}
var.diff <- ifelse(var.test(all.aml[i.mapk3,i.f],all.aml[i.mapk3,-i.f])$p.value < 0.05,TRUE,FALSE)
t.test(all.aml[i.mapk3,i.f],all.aml[i.mapk3,-i.f],alternative = "two.sided", var.equal = var.diff)
```

> Can we reject the H0 hypothesis?

------------------------------------------------------------------------

#### Exercise set C

This is the code needed to run the exercise without problem:

```{r}
all.aml = read.delim('http://bioinfo.ipmb.uni-heidelberg.de/crg/datascience3fs/practicals/data/all.aml.cleaned.csv',header=TRUE)
all.aml.anno = read.delim("http://bioinfo.ipmb.uni-heidelberg.de/crg/datascience3fs/practicals/data/all.aml.anno.cleaned.csv", header=TRUE)
#
# We convert all.aml into a data matrix rather than a data.frame
all.aml = data.matrix(all.aml)
#
i.mapk3 <- grep("MAPK3",row.names(all.aml))
i.ring1 <- grep("RING1",row.names(all.aml))
#
# Case 2:
boxplot(list(Female=all.aml[i.ring1,i.f],Male=all.aml[i.ring1,-i.f]),main=rownames(all.aml)[i.ring1])
#
# Case 3:
boxplot(list(BoneM=all.aml[i.mapk3,i.bm],P_Blood=all.aml[i.mapk3,-i.bm]),main=rownames(all.aml)[i.mapk3])
#
# Case 4:
boxplot(list(BoneM=all.aml[i.ring1,i.bm],P_Blood=all.aml[i.ring1,-i.bm]),main=rownames(all.aml)[i.ring1])
```

1)  For each case, formulate a H0 and H1 hypothesis and perform a t.test using the function t.test().

2)  For those genes that are normally distributed, can you perform a differential expression analysis (i.e. a t.test) between ALL and AML patients?

3)  Get the list of genes that are differentially expressed between ALL and AML.

## 4. Extra insight: building the H0 distribution

**Note: you can leave this part out...**

The `t-test` is based on a theoretical distribution under the H0 hypothesis.
We can however try to build this H0 distribution enpirically, using a sampling approach.

Let's compute the t-value for EIF4B for 2 **randomly chosen** groups within the cohort.
The rationale here: there is not reason why between 2 random groups, there should be a significant difference, right?

We create 2 random groups of patients, each one having the size of the ALL / AML group

```{r}
# Sample random columns 
i.randomALL  = sample(1:ncol(all.aml),length(i.all))
#
# Compute the t-test between the 2 random groups and get the t-value
t.h0=t.test(all.aml[i.gene,i.randomALL],
            all.aml[i.gene,-i.randomALL])$statistic
t.h0
```

> repeat this several times to get various t-values over various random groups We can repeat this 10000 times over 10000 random splits of the cohorts

```{r}
t.values = sapply(1:10000,function(i) {
  # random columns 
  i.randomALL  = sample(1:ncol(all.aml),length(i.all))
  #
  # get t-value between the 2 random groups
  t.test(all.aml[i.gene,i.randomALL],all.aml[i.gene,-i.randomALL])$statistic
})
#
# Plot the histogram
hist(t.values);abline(v=t.obs,lwd=3,col='red')
```

This distribution approximates the $H_0$ distribution of the test statistics.

> What does the distribution look like??
<<<<<<< Updated upstream:06_HypothesisTesting.Rmd
> Compute the mean and standard deviation of the t.values; plot a qq-plot
=======

We can compute the mean and standard deviation of the t.values and plot a qq-plot:
>>>>>>> Stashed changes:archive/06_HypothesisTesting.Rmd

```{r}
mean(t.values)
sd(t.values)
qqnorm(t.values);qqline(t.values)
```

This is a standard normal distribution (actually, a t-distribution with degrees of freedom (df) equal to the number of patients -2)!\
Do you remember why -2?

We can overlay the **empirical** $H_0$ distribution that we have just determined with the **theoretical** $H_0$ distribution, which is a t-distribution with `r ncol(all.aml)-2` degrees of freedom.

```{r}
x = seq(-4,4,by=0.2)
y = dt(x,df=ncol(all.aml)-2)
hist(t.values,freq=FALSE);lines(x,y,type='l',col= 'red',lwd=3)
```

> Pretty close, no?

## Computing the p-value

We can place the value of the observed t-statistics between the **real** ALL and AML subgroups on this histogram:

```{r}
hist(t.values,xlim=c(-4,4));abline(v=t.obs,lty=3,lwd=4,col='red')
```

Since we are performing a 2-sided t-test here (remember the $H_0$ hypothesis formulation!), we have to check what the probability is to get **more extreme** t-values under $H_0$.

<<<<<<< Updated upstream:06_HypothesisTesting.Rmd
> Check in how many of these 10000 random splits we get a t-value that is more extreme than the observed, hence : lower that -abs(t.obs) or larger than abs(t.obs).
> Normalize this number to the 10000 random splits to obtain the p-value!
=======
We can check in how many of these 10000 random splits we get a t-value that is more extreme than the observed, hence : lower than -abs(t.obs) or larger than abs(t.obs).
We then normalize this number to the 10000 random splits to obtain the p-value:
>>>>>>> Stashed changes:archive/06_HypothesisTesting.Rmd

```{r}
p.value = sum(t.values < -abs(t.obs) | t.values > abs(t.obs))/length(t.values)
p.value
```

Instead of using the empirical $H_0$ distribution, we can use the theoretical distribution (t-distribution):

```{r}
## Degrees of freedom df
df = ncol(all.aml)-2
##
## this is the one-sided p-value for the upper tail test
pt(t.obs, df = df, lower.tail=FALSE)
##
## this is the two-sided p-value
2*pt(t.obs, df = df, lower.tail=FALSE)
```

Finally, we can compare these computed p-value with the p-value returned by the function `t.test`:

```{r}
# If var.equal is not specified, the default is var.equal = FALSE. See help for the function t.test()
t.test(all.aml[i.gene,i.all],all.aml[i.gene,-i.all])
# You can extract the p-value using $p.value
t.test(all.aml[i.gene,i.all],all.aml[i.gene,-i.all])$p.value
#
# It is possible to evaluate the variance using the function var.test(). More about the underlying theory in the coming lectures!
var.test(all.aml[i.gene,i.all],all.aml[i.gene,-i.all])
# 
## Now extract the p-value
var.test(all.aml[i.gene,i.all],all.aml[i.gene,-i.all])$p.value
#
## This can be directly integrated into the t.test() using the function ifelse()
t.test(all.aml[i.gene,i.all],
       all.aml[i.gene,-i.all],
       var.equal = ifelse(var.test(all.aml[i.gene,i.all],all.aml[i.gene,-i.all])$p.value < 0.05,FALSE,TRUE))
```

> Do you consider this to be significant?

------------------------------------------------------------------------
